Solving Collective Spin Hamiltonians with Decoherence

### 2018-08-28 soc_squeezing/cumulant_expansions.tex
added -Z expectation value to correlator_methods.py; rearranged some text in expansions notes
added full, faithful explanation of how time-series correlators are computed, and add derivation of initial values for the state +X

\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\S}{\mathcal{S}}

@@

The case of the one-axis twisting is simplest to consider, as we can
write the Hamiltonian in the form
\begin{align}
  H_{\t{OAT}} = \f14 \chi \sum_{j,k} \sigma_j^z \sigma_k^z
  = \f12 \chi \sum_{j<k} \sigma_j^z \sigma_k^z + \f14 N \chi,
\end{align}
which admits an exact solution previously worked out by Foss-Feig et
al.~in ref.~\cite{foss-feig2013nonequilibrium}.  Adapting exact
expectation values with respect to the initial state $\ket{X}$
satisfying $S_x\ket{X}=S\ket{X}$, in terms of $\mu,\nu\in\set{+1,-1}$
we find that
\begin{align}
  \bk{S_+}
  &= \sum_j\bk{\sigma_j^+}
  = \f{N}{2} e^{-\gamma t/2} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_z}
  &= \f12\sum_j\bk{\sigma_j^\mu\sigma_j^z}
  + \f12\sum_{j\ne k} \bk{\sigma_j^\mu \sigma_k^z} \\
  &= -\f{\mu}{2}\bk{S_\mu} + \f14 N \p{N-1} e^{-\gamma t/2}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \sum_j \bk{\sigma_j^\mu \sigma_j^\nu}
  + \sum_{j\ne k} \bk{\sigma_j^\mu \sigma_k^\nu} \\
  &= \delta_{\mu,-\nu} \p{\f{N}{2} + \mu \bk{S_z}}
  + \f14 N \p{N-1} e^{-\gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{\xi,t} \equiv e^{-\gamma t/2}
  \sp{\cos\p{s_\xi t}+\f12\gamma t~\t{sinc}\p{s_\xi t}},
  &&
  \Psi\p{\xi,t} \equiv e^{-\gamma t/2}
  \p{is_\xi-\gamma/2}t~ \t{sinc}\p{s_\xi t},
\end{align}
and $s_\xi \equiv \xi + i2\gamma$.  The last ingredients we need
to compute spin squeezing at any time are $\bk{S_z}$ and $\bk{S_z^2}$.
As these operators commute with the OAT Hamiltonian, their evolution
is governed entirely by decoherence, such that
\begin{align}
  \f{d}{dt}\p{S_z^n}
  = \gamma \D_-\p{S_z^n}
  = \gamma \p{S + S_z} \sum_{k=0}^{n-1}
  \p{-1}^{n-k} { n \choose k } S_z^k,
\end{align}
where $S\equiv N/2$ and $\D_-\p{S_z^n}$ is worked out in Appendix
\ref{sec:decoherence}.  In particular,
\begin{align}
  \f{d}{dt} S_z = -\gamma \p{S + S_z},
  &&
  \f{d}{dt}\p{S_z^2} = \gamma \p{S + S_z} \p{1 - 2S_z}.
\end{align}
The initial conditions $\bk{S_z}_{t=0}=0$ and $\bk{S_z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_z} = S\p{e^{-\gamma t}-1},
  &&
  \bk{S_z^2} = S \p{1 - \f12 e^{-\gamma t}} e^{-\gamma t} + \bk{S_z}^2.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}-\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the squeezing parameter
$\xi^2$ in \eqref{eq:squeezing} at any time throughout evolution of
the initial state $\ket{X}:S_x\ket{X}=S\ket{X}$ under $H_{\t{OAT}}$.

\section{Cumulant expansion and Heisenberg systems of equations}

Unlike the one-axis twisting Hamiltonian, the two-axis twisting and
transverse-field Hamiltonians do not admit exact analytical
solutions. We will therefore solve these Hamiltonians approximately
using systems of equations for Heisenberg operators.  The basic idea
is to find equations of motion for operators of interest, and at some
point truncate high-order terms to arrive at a closed,
reasonably-sized system of equations.  To benchmark this method, we
will test it against exact one-axis twisting results.

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_z^q S_-^r.
  \label{eq:H_general}
\end{align}
The Hamiltonians for one-axis twisting with a transverse-field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_z^2 + \Omega S_x
  = \chi S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{z,y}
  = \chi \p{S_z^2 - S_y^2}
  = \chi \p{S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z},
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{y,x}
  = \chi\p{S_y^2 - S_x^2}
  = -\f{\chi}{2} \p{S_+^2 + S_-^2}.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_z^m S_-^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_z^q S_-^r, S_+^\ell S_z^m S_-^n}_-
  + \gamma \D_-\p{S_+^\ell S_z^m S_-^n}
  \equiv \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_z^q S_-^r,
\end{align}
where the commutator is worked out in Appendix \ref{sec:comm_general},
and the decoherence operator in Appendix \ref{sec:decoherence}.  We
can then expand
\begin{align}
  \bk{S_+^\ell S_z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_z^m S_-^n}}_{t=0}
  \approx \sum_{k=0}^{M-1} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_z^m S_-^n}}_{t=0},
\end{align}
where $M$ is chosen such that
\begin{align}
  \f{t^M}{M!} \bk{\f{d^M}{dt^M}\p{S_+^\ell S_z^m S_-^n}}_{t=0}
  \ll S^{\ell+m+n}.
\end{align}
Initial conditions, i.e.~expectation values at $t=0$, are provided in
Appendix \ref{sec:initial_conditions}.
% comment on explosion of operators

@@

Initial conditions for all collective spin operators can be determined
from the expansions
\begin{align}
  S_z^n = \sum_{k=-S}^S k^n \op{k},
  &&
  S_\mu
  = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{k,1}}
  g_\mu\p{k} \op{k+\mu}{k},
\end{align}
with
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}},
\end{align}
and the initial states
\begin{align}
  \ket{X} \equiv 2^{-N/2} \sum_{k=-S}^S { N \choose S+k }^{1/2} \ket{k},
  &&
  \ket{-Z} \equiv \ket{-S},
\end{align}
using which we can compute
\begin{align}
  \bk{X|S_+^\ell S_z^m S_-^n|X}
  &=  2^{-N} \sum_{k=-S}^{S-\max\set{\ell,n}} k^m
  \sp{{ N \choose S+k+\ell } { N \choose S+k+n }}^{1/2} \tag*{} \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k+p}
  \prod_{q=0}^{n-1} g_-\p{S,k+n-q} \\
  &= 2^{-N} \sum_{k=-S}^{S-\max\set{\ell,n}} k^m
  \f{N!\p{S-k}!}{\p{S+k}!\p{S-k-\ell}!\p{S-k-n}!}.
\end{align}
and
\begin{align}
  \bk{-Z|S_+^\ell S_z^m S_-^n|-Z}
  = \delta_{\ell,0} \delta_{n,0} \p{-S}^m.
\end{align}

>>>>

\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\S}{\mathcal{S}}

@@

The case of the one-axis twisting is simplest to consider, as we can
write the Hamiltonian in the form
\begin{align}
  H_{\t{OAT}} = \f14 \chi \sum_{j,k} \sigma_j^z \sigma_k^z
  = \f12 \chi \sum_{j<k} \sigma_j^z \sigma_k^z + \f14 N \chi,
\end{align}
which admits an exact solution previously worked out by Foss-Feig et
al.~in ref.~\cite{foss-feig2013nonequilibrium}.  Adapting exact
expectation values with respect to the initial state $\ket{X}$
satisfying $S_x\ket{X}=S\ket{X}$, in terms of $\mu,\nu\in\set{+1,-1}$
we find that
\begin{align}
  \bk{S_+}
  &= \sum_j\bk{\sigma_j^+}
  = \f{N}{2} e^{-\gamma t/2} \Phi\p{\chi,t}^{N-1}, \label{eq:S+_OAT} \\
  \bk{S_\mu S_z}
  &= \f12\sum_j\bk{\sigma_j^\mu\sigma_j^z}
  + \f12\sum_{j\ne k} \bk{\sigma_j^\mu \sigma_k^z} \\
  &= -\f{\mu}{2}\bk{S_\mu} + \f14 N \p{N-1} e^{-\gamma t/2}
  \Psi\p{\mu\chi,t} \Phi\p{\chi,t}^{N-2}, \\
  \bk{S_\mu S_\nu}
  &= \sum_j \bk{\sigma_j^\mu \sigma_j^\nu}
  + \sum_{j\ne k} \bk{\sigma_j^\mu \sigma_k^\nu} \\
  &= \delta_{\mu,-\nu} \p{\f{N}{2} + \mu \bk{S_z}}
  + \f14 N \p{N-1} e^{-\gamma t}
  \Phi\p{\sp{\mu+\nu}\chi,t}^{N-2}, \label{eq:SS+-_OAT}
\end{align}
where
\begin{align}
  \Phi\p{\xi,t} \equiv e^{-\gamma t/2}
  \sp{\cos\p{s_\xi t}+\f12\gamma t~\t{sinc}\p{s_\xi t}},
  &&
  \Psi\p{\xi,t} \equiv e^{-\gamma t/2}
  \p{is_\xi-\gamma/2}t~ \t{sinc}\p{s_\xi t},
\end{align}
and $s_\xi \equiv \xi + i2\gamma$.  The last ingredients we need
to compute spin squeezing at any time are $\bk{S_z}$ and $\bk{S_z^2}$.
As these operators commute with the OAT Hamiltonian, their evolution
is governed entirely by decoherence, such that
\begin{align}
  \f{d}{dt}\p{S_z^n}
  = \gamma \D_-\p{S_z^n}
  = \gamma \p{S + S_z} \sum_{k=0}^{n-1}
  \p{-1}^{n-k} { n \choose k } S_z^k,
\end{align}
where $S\equiv N/2$ throughout this work and $\D_-\p{S_z^n}$ is worked
out in Appendix \ref{sec:decoherence}.  In particular,
\begin{align}
  \f{d}{dt} S_z = -\gamma \p{S + S_z},
  &&
  \f{d}{dt}\p{S_z^2} = \gamma \p{S + S_z} \p{1 - 2S_z}.
\end{align}
The initial conditions $\bk{S_z}_{t=0}=0$ and $\bk{S_z^2}_{t=0}=S/2$
then imply
\begin{align}
  \bk{S_z} = S\p{e^{-\gamma t}-1},
  &&
  \bk{S_z^2} = S \p{1 - \f12 e^{-\gamma t}} e^{-\gamma t} + \bk{S_z}^2.
  \label{eq:Sz_OAT}
\end{align}
The expectation values in \eqref{eq:S+_OAT}-\eqref{eq:SS+-_OAT} and
\eqref{eq:Sz_OAT} are sufficient to compute the squeezing parameter
$\xi^2$ in \eqref{eq:squeezing} at any time throughout evolution of
the initial state $\ket{X}:S_x\ket{X}=S\ket{X}$ under $H_{\t{OAT}}$.

\section{Cumulant expansion and systems of equations for Heisenberg
  operators}

Unlike the one-axis twisting Hamiltonian, the two-axis twisting and
transverse-field Hamiltonians do not admit exact analytical
solutions. We will therefore solve these Hamiltonians approximately
using systems of equations for Heisenberg operators.  The basic idea
is to find equations of motion for operators of interest, and at some
point truncate high-order terms with negligible contribution to the
spin correlators of interest.  To benchmark this method, we test it
against exact one-axis twisting results in the presence of
decoherence, as well as transverse-field and two-axis twisting in the
absence of decoherence.

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_z^q S_-^r,
  \label{eq:H_general}
\end{align}
where Hermiticity requires that $h_{pqr}=h_{rqp}$.  The Hamiltonians
for one-axis twisting with a transverse field, for example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_z^2 + \Omega S_x
  = \chi S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{z,y}
  = \chi \p{S_z^2 - S_y^2}
  = \chi \p{S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z},
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{y,x}
  = \chi\p{S_y^2 - S_x^2}
  = -\f{\chi}{2} \p{S_+^2 + S_-^2}.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_z^m S_-^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_z^q S_-^r, S_+^\ell S_z^m S_-^n}_-
  + \gamma \D_-\p{S_+^\ell S_z^m S_-^n}
  \equiv \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_z^q S_-^r.
\end{align}
Equipped with a matrix representation of the time derivative operator,
i.e.~$T^{\ell m n}_{pqr}$, we can expand
\begin{align}
  \bk{S_+^\ell S_z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_z^m S_-^n}}_{t=0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{p,q,r} \sp{T^k}^{\ell m n}_{pqr} \bk{S_+^p S_z^q S_-^r}_{t=0},
  \label{eq:time_series}
\end{align}
where
\begin{align}
  \sp{T^k}^{\ell mn}_{pqr}
  \equiv \sum_{\v\lambda,\v\mu,\v\nu}
  T^{\ell mn}_{\lambda_1 \mu_1 \nu_1}
  T^{\lambda_1 \mu_1 \nu_1}_{\lambda_2 \mu_2 \nu_2}
  T^{\lambda_2 \mu_2 \nu_2}_{\lambda_3 \mu_3 \nu_3}
  \cdots T^{\lambda_{k-1} \mu_{k-1} \nu_{k-1}}_{pqr}.
\end{align}
As terms in \eqref{eq:time_series} are suppressed factorially in $k$,
for any fixed (or maximal) time $t$ we can truncate the series to
write
\begin{align}
  \bk{S_+^\ell S_z^m S_-^n}
  \approx \sum_{k=0}^{M_t-1} \f{t^k}{k!}
  \sum_{p,q,r} \sp{T^k}^{\ell m n}_{pqr} \bk{S_+^p S_z^q S_-^r}_{t=0},
\end{align}
where $M_t$ is chosen such that
\begin{align}
  \f{t^{M_t}}{M_t!} \sum_{p,q,r}
  \sp{T^{M_t}}^{\ell m n}_{pqr} \bk{S_+^p S_z^q S_-^r}_{t=0}
  \ll S^{\ell+m+n}.
\end{align}
The commutator and decoherence operator necessary for computing the
time derivative operator $T^{\ell m n}_{pqr}$ are respectively worked
out in Appendix \ref{sec:comm_general} and \ref{sec:decoherence},
while expectation values for initial conditions of interest are
provided in Appendix \ref{sec:initial_conditions}.
% comment on explosion of operators

@@

Initial conditions for all collective spin operators can be determined
from the expansions
\begin{align}
  S_z^n = \sum_{k=-S}^S k^n \op{k},
  &&
  S_\mu
  = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{k,1}}
  g_\mu\p{k} \op{k+\mu}{k},
\end{align}
with
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}},
\end{align}
and the initial states
\begin{align}
  \ket{-Z} \equiv \ket{-S},
  &&
  \ket{X} \equiv 2^{-N/2} \sum_{k=-S}^S { N \choose S+k }^{1/2} \ket{k},
\end{align}
using which we can compute
\begin{align}
  \bk{-Z|S_+^\ell S_z^m S_-^n|-Z}
  = \delta_{\ell,0} \delta_{m,0} \p{-S}^m,
\end{align}
and
\begin{align}
  \bk{X|S_+^\ell S_z^m S_-^n|X}
  &=  2^{-N} \sum_{k=-S}^{S-\max\set{\ell,n}} k^m
  \sp{{ N \choose S+k+\ell } { N \choose S+k+n }}^{1/2} \tag*{} \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k+p}
  \prod_{q=0}^{n-1} g_-\p{S,k+n-q} \\
  &= \f{N!}{2^N} \sum_{k=-S}^{S-\max\set{\ell,n}} k^m
  \f{\p{S-k}!}{\p{S+k}!\p{S-k-\ell}!\p{S-k-n}!}.
\end{align}
With the help of Mathematica, we can evaluate the sum to find
\begin{align}
  \bk{X|S_+^\ell S_z^m S_-^n|X}
  = \f{\p{N!}^2}{2^N\p{N-\ell}!\p{N-n}!}
  \sp{\p{-S}^m \alpha + \f{\p{N-\ell}\p{N-n}}{N} \beta},
\end{align}
where $\alpha$ and $\beta$ are defined in terms of the generalized
hypergeometric function $\F$ by
\begin{align}
  \alpha
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{\ell-N,~n-N\\-N}~; -1},
\end{align}
and
\begin{align}
  \beta
  \equiv \sum_{p=0}^{m-1} \p{-S}^p { m \choose p }
  \tilde\F_{m-p-1}\p{\let\scriptstyle\textstyle
    \substack{1+\ell-N,~1+n-N\\1-N}~; -1},
\end{align}
for
\begin{align}
  \tilde\F_q\p{\let\scriptstyle\textstyle
    \substack{a,~b\\c}~; z}
  \equiv \F\p{\let\scriptstyle\textstyle
    \substack{2,2,\cdots,2,a,b\\1,1,\cdots,1,c}~; z}
\end{align}
with $q$ repetitions of 2 and 1.
